---
title: "Midterm 1"
author: "Carmina Inguito"
date: "`2022-01-27`"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your code should be organized, clean, and run free from errors. Remember, you must remove the `#` for any included code chunks to run. Be sure to add your name to the author header above. You may use any resources to answer these questions (including each other), but you may not post questions to Open Stacks or external help sites. There are 15 total questions, each is worth 2 points.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

This exam is due by 12:00p on Thursday, January 27.  

## Load the tidyverse
If you plan to use any other libraries to complete this assignment then you should load them here.
```{r}
library(tidyverse)
library("janitor")
library(skimr)
```

## Questions  
Wikipedia's definition of [data science](https://en.wikipedia.org/wiki/Data_science): "Data science is an interdisciplinary field that uses scientific methods, processes, algorithms and systems to extract knowledge and insights from noisy, structured and unstructured data, and apply knowledge and actionable insights from data across a broad range of application domains."  

**1. (2 points) Consider the definition of data science above. Although we are only part-way through the quarter, what specific elements of data science do you feel we have practiced? Provide at least one specific example.**  
In these past few weeks, we have practiced various elements of data science including manipulating variables, creating data frames, making simple computations, and overall learning how to organize data in a more straightforward way that would make sense to any individual who comes across it (no matter if they have/haven't had previous knowledge of the program). For instance, removing NAs (missing data) within the data by using `na.rm=T`. We can check beforehand for missing data by utilizing `is.na`to help deal with it and keep data neat. While we have removed multiple NAs throughout the quarter so far, one can refer to Lab 4 in regards to organizing herbivore/carnivore data. 

**2. (2 points) What is the most helpful or interesting thing you have learned so far in BIS 15L? What is something that you think needs more work or practice?**  
As I have experienced using RStudio in previous classes like STA100 and my MAT17A class at UC Davis, it was really helpful to be able to delve deeper into the organization of different data and being able to ultimately pull out a specific piece of information you're looking for. From being able to get sort of a refresher on how RStudio works to now being able to create new objects, make data frames, and explore new functions- it certainly has been a hands-on learning experience. I thought creating our own repository and navigating through GitHub was also an interesting first for me. 

Something that I think I need more practice on is getting used to utilizing pipes `%>%`, `if else` or the `|` function. 

In the midterm 1 folder there is a second folder called `data`. Inside the `data` folder, there is a .csv file called `ElephantsMF`. These data are from Phyllis Lee, Stirling University, and are related to Lee, P., et al. (2013), "Enduring consequences of early experiences: 40-year effects on survival and success among African elephants (Loxodonta africana)," Biology Letters, 9: 20130011. [kaggle](https://www.kaggle.com/mostafaelseidy/elephantsmf).  

**3. (2 points) Please load these data as a new object called `elephants`. Use the function(s) of your choice to get an idea of the structure of the data. Be sure to show the class of each variable.**

```{r}
elephants <- readr::read_csv("data/ElephantsMF.csv") #Create a new object 'elephants' and load the data
```

```{r}
glimpse(elephants) #Summary of data
```

```{r}
skim(elephants) #Another perspective of the data
```

**4. (2 points) Change the names of the variables to lower case and change the class of the variable `sex` to a factor.**
```{r}
elephants <- janitor::clean_names(elephants) #Removing an uppercase words by using the Janitor package
names(elephants)
```
```{r}
elephants <- elephants %>% #Changing class of variable 'sex' to a factor
  mutate(sex=as.factor(sex))
```

```{r}
class(elephants$sex)
```

**5. (2 points) How many male and female elephants are represented in the data?**
```{r}
table(elephants$sex) 
```

There are 138 males and 150 female elephants represented in the data.

**6. (2 points) What is the average age all elephants in the data?**
```{r}
mean(elephants$age, na.rm=T) 
```
The average age of all elephants in the data is about 10.97 or 11 if rounded.

**7. (2 points) How does the average age and height of elephants compare by sex?**

```{r}
elephants %>%
  group_by(sex) %>%
  summarise(average_age=mean(age),
            average_height=mean(height))
```
Females have a greater average height and age compared to the males. 

**8. (2 points) How does the average height of elephants compare by sex for individuals over 20 years old. Include the min and max height as well as the number of individuals in the sample as part of your analysis.**  

```{r}
elephants %>%
  filter(age>20) %>%
  group_by(sex) %>%
  summarise(mean_height=mean(height),
            max_height=max(height), 
            min_height=min(height),
            n_elephants=n())
```
According to the data, females that are over 20 years old have an average height of 232.2 while males over 20 years old have an average height of 269.6 making them with the higher average. 

For the next series of questions, we will use data from a study on vertebrate community composition and impacts from defaunation in [Gabon, Africa](https://en.wikipedia.org/wiki/Gabon). One thing to notice is that the data include 24 separate transects. Each transect represents a path through different forest management areas.  

Reference: Koerner SE, Poulsen JR, Blanchard EJ, Okouyi J, Clark CJ. Vertebrate community composition and diversity declines along a defaunation gradient radiating from rural villages in Gabon. _Journal of Applied Ecology_. 2016. This paper, along with a description of the variables is included inside the midterm 1 folder.  

**9. (2 points) Load `IvindoData_DryadVersion.csv` and use the function(s) of your choice to get an idea of the overall structure. Change the variables `HuntCat` and `LandUse` to factors.**

```{r}
gabon_africa <-readr::read_csv("data/IvindoData_DryadVersion.csv")
```

```{r}
summary(gabon_africa) #Getting an overview of the data
```

```{r}
gabon_africa <- janitor::clean_names(gabon_africa)
names(gabon_africa)
```

```{r}
gabon_africa <- gabon_africa %>% #All names to lowercase
  mutate_all(tolower)
gabon_africa
```

```{r}
gabon_africa <- gabon_africa %>%
  mutate(land_use=as.factor(land_use), hunt_cat=as.factor(hunt_cat)) #Changing hunt_cat and land_use to factor
```

```{r}
class(gabon_africa$land_use) #Checking what the variable identifies as 
```
```{r}
class(gabon_africa$hunt_cat) #Checking what the variable identifies as 
```

**10. (4 points) For the transects with high and moderate hunting intensity, how does the average diversity of birds and mammals compare?**

```{r}
levels(gabon_africa$hunt_cat) #Ensuring the hunt_cat levels are spelled correctly and lowercase
```
```{r}
gabon_africa$diversity_bird_species <- as.numeric(gabon_africa$diversity_bird_species)
class(gabon_africa$diversity_bird_species) #Checking the change from factor to numeric
```
```{r}
gabon_africa$diversity_mammal_species <- as.numeric(gabon_africa$diversity_mammal_species)
class(gabon_africa$diversity_mammal_species) #Checking the change from factor to numeric
```

```{r}
gabon_africa %>%
  filter(hunt_cat == "high" | hunt_cat == "moderate") %>% 
  summarise(avg_bird_diversity=mean(diversity_bird_species),avg_mammal_diversity=mean(diversity_mammal_species)) #Calculating the mean diversity
```
The average diversity between the birds and the mammals are actually fairly close. Mammals have a slightly greater diversity mean at approximately 1.71 while the birds have a diversity mean of approximately 1.64.

**11. (4 points) One of the conclusions in the study is that the relative abundance of animals drops off the closer you get to a village. Let's try to reconstruct this (without the statistics). How does the relative abundance (RA) of apes, birds, elephants, monkeys, rodents, and ungulates compare between sites that are less than 3km from a village to sites that are greater than 25km from a village? The variable `Distance` measures the distance of the transect from the nearest village. Hint: try using the `across` operator.**  

```{r}
names(gabon_africa)
```

```{r}
gabon_africa$distance <- as.numeric(gabon_africa$distance) #factor to numeric
gabon_africa$ra_ungulate <- as.numeric(gabon_africa$ra_ungulate)
gabon_africa$ra_rodent <- as.numeric(gabon_africa$ra_rodent)
gabon_africa$ra_monkeys <- as.numeric(gabon_africa$ra_monkeys)
gabon_africa$ra_elephant <- as.numeric(gabon_africa$ra_elephant)
gabon_africa$ra_birds <- as.numeric(gabon_africa$ra_birds)
gabon_africa$ra_apes <- as.numeric(gabon_africa$ra_apes)
```

```{r}
gabon_africa %>%
  filter(distance<3) %>% #filtering for distance under 3km
  summarise((across(c(ra_apes, ra_birds, ra_elephant, ra_monkeys, ra_rodent, ra_ungulate), mean, na.rm=T)))
```
```{r}
gabon_africa %>%
  filter(distance>25) %>% #filtering for distance over 25km 
  summarise((across(c(ra_apes, ra_birds, ra_elephant, ra_monkeys, ra_rodent, ra_ungulate), mean, na.rm=T)))
```

**12. (4 points) Based on your interest, do one exploratory analysis on the `gabon` data of your choice. This analysis needs to include a minimum of two functions in `dplyr.`**

```{r}
levels(gabon_africa$land_use)
```

Here, I am interested in finding the greatest amount (max) of species, least (min) amount of mammal species and the mean-- who utilize the park as their land.
```{r}
gabon_africa %>%
  filter(land_use=="park") %>%
  group_by(diversity_mammal_species) %>%
  summarise(avg_mammal_all=mean(diversity_mammal_species),min_mammal_all=min(diversity_mammal_species), max_mammal_all=max(diversity_mammal_species))
```

