---
title: "BIS 15L Midterm 2"
author: "Carmina Inguito"
date: "`2022-17-02`"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your code should be organized, clean, and run free from errors. Be sure to **add your name** to the author header above. You may use any resources to answer these questions (including each other), but you may not post questions to Open Stacks or external help sites. There are 10 total questions.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean! Your plots should use consistent aesthetics throughout.  

This exam is due by **12:00p on Tuesday, February 22**.  

## Gapminder
For this assignment, we are going to use data from  [gapminder](https://www.gapminder.org/). Gapminder includes information about economics, population, social issues, and life expectancy from countries all over the world. We will use three data sets, so please load all three as separate objects.  

```{r}
library("ggthemes")
library("gapminder")
library("skimr")
library("tidyverse")
library("here")
```


1. population_total.csv  
```{r}
population_total <-readr::read_csv(here("midterm2", "data", "population_total.csv")) 
```

2. income_per_person_gdppercapita_ppp_inflation_adjusted.csv  
```{r}
income_per_person <-readr::read_csv(here("midterm2", "data", 
"income_per_person_gdppercapita_ppp_inflation_adjusted.csv"))
```

3. life_expectancy_years.csv  
```{r}
life_expectancy <- readr::read_csv(here("midterm2", "data", "life_expectancy_years.csv"))
```

**1. (3 points) Once you have an idea of the structure of the data, please make each data set tidy (hint: think back to pivots) and store them as new objects. You will need both the original (wide) and tidy (long) data!**

**Population Total**
```{r}
anyNA(population_total)
```

```{r}
head(population_total)
```

```{r}
long_population <- population_total %>%
  pivot_longer(-c(country),
               names_to="year",
               values_to="population")
long_population
```

**Income**
```{r}
anyNA(income_per_person)
```

```{r}
head(income_per_person)
```

```{r}
long_income <- income_per_person %>%
  pivot_longer(-c(country),
               names_to="year",
               values_to="gdp_per_capita")
long_income
```

**Life Expectancy**
```{r}
anyNA(life_expectancy)
```

```{r}
head(life_expectancy)
```

```{r}
long_life_exp <- life_expectancy %>%
  pivot_longer(-c(country),
               names_to= "year",
               values_to= "life_expectancy")
long_life_exp
```

**2. (1 point) How many different countries are represented in the data? Provide the total number and their names. Since each data set includes different numbers of countries, you will need to do this for each one.**  

**Population**
```{r}
long_population %>%
  summarise(n=n_distinct(country)) %>%
  summarise(country_count=sum(n))
```
```{r}
long_population %>%
  distinct(country)
```

There are 195 different countries represented in the data.

**Income**
```{r}
long_income %>%
  summarise(n=n_distinct(country)) %>%
  summarise(country_count=sum(n))
```

```{r}
long_income %>%
  distinct(country)
```
There are 193 distinct countries within the income per person data.

**Life Expectancy**
```{r}
long_life_exp %>%
  summarise(n=n_distinct(country)) %>%
  summarise(country_count=sum(n))
```

```{r}
long_life_exp %>%
  distinct(country)
```

There are 187 distinct countries within the life expectancy data.

```{r}
long_life_exp
```


## Life Expectancy  

**3. (2 points) Let's limit the data to 100 years (1920-2020). For these years, which country has the highest average life expectancy? How about the lowest average life expectancy?** 

**Highest Average Life Expectancy** Andorra
```{r}
long_life_exp %>%
  filter(year <=2020 & year>=1920) %>% 
  group_by(country) %>%
  summarise(mean_life_expectancy=mean(life_expectancy, na.rm=T)) %>%
  arrange(desc(mean_life_expectancy)) %>%
  head(n=1)
```

**Lowest Average Life Expectancy** Central African Republic
```{r}
long_life_exp %>%
  filter(year <=2020 & year>=1920) %>% 
  group_by(country) %>%
  summarise(mean_life_expectancy=mean(life_expectancy, na.rm=T)) %>%
  arrange(mean_life_expectancy) %>%
  head(n=1)
```

**4. (3 points) Although we can see which country has the highest life expectancy for the past 100 years, we don't know which countries have changed the most. What are the top 5 countries that have experienced the biggest improvement in life expectancy between 1920-2020?**  
```{r}
long_life_exp %>% 
  filter(year==1920 | year==2020) %>%
  group_by(country) %>%
  summarise(life_exp_change=(max(life_expectancy)-min(life_expectancy))) %>%
  arrange(desc(life_exp_change)) %>%
  head(5)
```

The top five countries that have experienced the biggest improvement in life expectancy between 1920-2020 are Kuwait, Kyrgyz Republic, Turkmenistan, South Korea,and Tajikistan. 

**5. (3 points) Make a plot that shows the change over the past 100 years for the country with the biggest improvement in life expectancy. Be sure to add appropriate aesthetics to make the plot clean and clear. Once you have made the plot, do a little internet searching and see if you can discover what historical event may have contributed to this remarkable change.**

```{r}
long_life_exp %>%
  filter(country== "Kuwait",
         !is.na(life_expectancy), 
         between(year, 1920,2020)) %>%

ggplot(aes(x=as.numeric(year), y=life_expectancy, color=country))+
  geom_line() +
  labs(x="Year", 
       y="Life Expectancy (years)",
       title="Kuwait's Change in Life Expectancy From 1920-2020",
       color="Country") +
  theme_light()+
theme(plot.title=element_text(face="bold", hjust=0.5))
```

As one can see from the created plot, Kuwait has greatly increased in life expectancy over a duration of 100 years. Around the 1930s, the Petroleum Industry was created which paved way for the expansion in the economy; thus, allowing the country to be more wealthier and have a greater longevity for those living in Kuwait. 

## Population Growth
**6. (3 points) Which 5 countries have had the highest population growth over the past 100 years (1920-2020)?**
```{r}
options(scipen=100) #did research here as I wasn't sure how to remove the scientific notation
long_population %>%
  filter(!is.na(population), year %in% c(1920,2020)) %>%
  group_by(country) %>%
  summarise(population_growth=(max(population)-min(population))) %>%
  arrange(desc(population_growth)) %>%
  head(5)
```

The top five countries with the highest population growth over the past 100 years are India, China, Indonesia,United States, and Pakistan. 

**7. (4 points) Produce a plot that shows the 5 countries that have had the highest population growth over the past 100 years (1920-2020). Which countries appear to have had exponential growth?**  
```{r}
palette <- c("#46AED4", "#4681D4", "#D47146", "#D44657",
          "#BF46D4") #interested in using different colors
```

```{r}
long_population %>%
  filter(year>=1920 & year<=2020) %>%
  filter(country==c("India", "China", "Indonesia", "United States", "Pakistan")) %>% 
  ggplot(aes(x=as.numeric(year), y=population, color=country))+
  geom_line(size=0.60)+
  labs(title= "Top Five Countries in Population Growth from 1920-2020",
       x= "Year",
       y= "Population (Individuals)")+
  theme_light()+
  scale_color_manual(values=palette)+
  theme(plot.title=element_text(face="bold",hjust=0.5))
long_population
```
The countries that appear to have exponential growth are **China and India.**

## Income
The units used for income are gross domestic product per person adjusted for differences in purchasing power in international dollars.

**8. (4 points) As in the previous questions, which countries have experienced the biggest growth in per person GDP. Show this as a table and then plot the changes for the top 5 countries. With a bit of research, you should be able to explain the dramatic downturns of the wealthiest economies that occurred during the 1980's.**  

**Note:** Change in GDP between countries- end/beginning year (doesn't account for good and bad years) subtract beginning and end? max/min? Results may differ depending on how you go about this. 

```{r}
years <- c(1920:2020)
colnames(long_income)
```

```{r}
income_overtime <- long_income %>% 
  mutate(year = str_replace(year, "x", ""),
         year = as.numeric(year)) %>%
  filter(year %in% years) %>% 
  pivot_wider(names_from = "year", values_from = "gdp_per_capita") %>% 
  mutate(change = `2020` - `1920`) %>% 
  select(c(country, change)) %>% 
  arrange(desc(change)) 
```

*The top five countries are Qatar, Luxembourg, Brunei, Singapore, and Ireland. Here's how each of their income has changed over 100 years (1920-2020):*

```{r}
long_income %>%
filter(country==c("Ireland", "Brunei", "Qatar", "Singapore", "Luxembourg")) %>%
  filter(year >=1920, year <=2020) %>%
  ggplot(aes(x=as.numeric(year),y=gdp_per_capita, fill=country, color=country)) +
  geom_line(size=0.60)+
  scale_color_manual(values=palette)+
  labs(title= "Top 5 Countries with Highest GDP Per Capita (1920-2020)", x="Year", y="GDP Per Capita") +
  theme_light()+
  theme(plot.title=element_text(face="bold",hjust=0.5))
  
```

During the 1980s, these wealthy economies had "dramatic downturns" because of the 1979 Oil Crisis as oil prices were decreasing internationally. Qatar, the most country that demonstrated the greatest decrease in GDP Per Capita in the plot, had its government's spending cut in order to help lower income. 

**9. (3 points) Create three new objects that restrict each data set (life expectancy, population, income) to the years 1920-2020. Hint: I suggest doing this with the long form of your data. Once this is done, merge all three data sets using the code I provide below. You may need to adjust the code depending on how you have named your objects. I called mine `life_expectancy_100`, `population_100`, and `income_100`. For some of you, learning these `joins` will be important for your project.**  
```{r}
life_expectancy_100 <- long_life_exp %>%
  filter(between(year, 1920, 2020))
```

```{r}
population_100 <- long_population %>%
  filter(between(year, 1920, 2020))
```

```{r}
income_100 <- long_income %>%
  filter(between(year, 1920, 2020))
```

```{r}
gapminder_join <- inner_join(life_expectancy_100, population_100, by= c("country", "year"))
gapminder_join <- inner_join(gapminder_join, income_100, by= c("country", "year"))
gapminder_join
```

**10. (4 points) Use the joined data to perform an analysis of your choice. The analysis should include a comparison between two or more of the variables `life_expectancy`, `population`, or `income.`**  

*I am interested in understanding the relationship between life_expectancy and income (GDP) between countries. If there's lower GDP, would that mean there's a decrease in life expectancy? What are alternative explanations as to why this may or may not be the case?*

```{r}
gapminder_join %>%
  group_by(country) %>%
  summarise(min=min(gdp_per_capita),
            max=max(gdp_per_capita)) %>%
  mutate(delta_income=max-min) %>%
  arrange(delta_income) %>%
  head(n=6)
```
Plotting the Data Above: 
```{r}
gapminder_join %>%
  filter(country==c("Mozambique", "Sierra Leone", "Malawi", "Niger", "Somalia", "Burundi")) %>%
  ggplot(aes(x=as.numeric(gdp_per_capita), y=life_expectancy))+ geom_point(size=1.25)+
  geom_smooth(method="lm", se=FALSE, color="purple")+
  facet_wrap(.~country)+
  theme_gray()+
  labs(title= "Comparing Life Expectancy & GDP Per Capita",
       x="GDP Per Individual",
       y="Life Expectancy (years)")+
  theme(plot.title=element_text(face="bold",hjust=0.5))
```

Interestingly enough, there were some countries that had a decrease in life expectancy as GDP increased. This could possibly be interpreted by income inequality in such places like Somalia and Niger. Higher per capita doesn't always necessarily mean a greater longevity. This may also be interpreted in a way in which a higher GDP leads to unhealthier habits (drugs, obesity due to greater access to food, or alcohol) that decrease life expectancy.